---
title: "ProteinDigestibilityAnalysisRMarkdown"
author: "Wendy Liu"
date: "2026-08-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)    # load the package
library(ggplot2)
library(dplyr)

```

## Calculating Digestibility Coefficients



```{r Transforming Data}

DigestibilityData <- read.csv("/Users/wendy/Downloads/Stirling_Digestibility.csv")

Controls = subset(DigestibilityData, DigestibilityData$Type == "Control")

Samples = subset(DigestibilityData, DigestibilityData$Type == "Sample")

# Calculate average Protein_Kjeldahl per Sample.Name in Controls
control_avg = aggregate(Protein_Kjeldahl ~ Sample.Name, data = Controls, FUN = mean)
names(control_avg)[2] = "averageControl"

# Merge that average into the Samples dataframe, matching by Sample.Name
Samples = merge(Samples, control_avg, by = "Sample.Name")

Samples$DigestibilityCoeff = 1- (Samples$Protein_Kjeldahl/Samples$averageControl)



```
## Statistics

Question: Does the ingredient of the feed (Sample.Name) has a significant impact on the digestibility coefficient (DigestibilityCoeff) ? 

DigestibilityCoeff ~ Sample.Name

Digestibility coefficient is measured independently i.e. only after the HG phase; 
It is also a continuous proportions (not from discrete counts), especially with values clustering near 0 or 1.

The first thing we need to check is the normality of the data contribution

```{r Normality test if we are to do ANOVA}

range(Samples$DigestibilityCoeff, na.rm = TRUE)
# This confirmed the proportions/percentages are indeed bounded 0-1

res.aov <- aov(DigestibilityCoeff ~ Sample.Name, data = Samples) #ANOVA

plot(res.aov, 2) #QQplot
aov_residuals <- residuals(object = res.aov) #extract ANOVA residuals
shapiro.test(x = aov_residuals) #run shapiro-wilk test

#  p < 0.05 -> residuals significantly deviate from normality


Samples$asin_DC <- asin(sqrt(Samples$DigestibilityCoeff))

res.aov <- aov(asin_DC ~ Sample.Name, data = Samples) #ANOVA
aov_residuals <- residuals(object = res.aov) #extract ANOVA residuals
shapiro.test(x = aov_residuals) #run shapiro-wilk test

#  p < 0.05 -> residuals significantly deviate from normality even after data transformation

```

Since our proportional data does not mean the assumptions for a one-way anlaysis of varianace, we can either:

1. Use beta regression. It's built specifically for outcomes bounded in (0,1) and doesn't assume normality. It will give us a test of whether your input variable significantly predicts the proportion.

2. non-parametric comparison between groups
The Kruskal-Wallis test Do the distributions of several independent groups differ? Rather than the analysis of variance (is there a difference in mean), 
(a) Null hypothesis: The outcome has the same distribution in all groups.
(b) Alternative hypothesis:The outcome distribution differs for at least one group.

Mann-Whitney U test (2 groups) or Kruskal-Wallis test (3+ groups) — these compare rank distributions and don't assume normality. Simple and often sufficient if you just want a "is there a difference" answer without modeling mechanics.

## 1. Beta regression

```{r Beta regression: batch effect}
library(glmmTMB)
library(lmtest)

# we can also test whether there is a batch effect the Date: 
model_1 <- glmmTMB(DigestibilityCoeff ~ Sample.Name,
                 data = Samples,
                 family = beta_family(link = "logit"))

Samples$Date <- as.factor(Samples$Date)

model_2 <- glmmTMB(DigestibilityCoeff ~ Sample.Name + (1| Date),
                 data = Samples,
                 family = beta_family(link = "logit"))

lrtest(model_1, model_2)

```

As p > 0.05, the liklyhood ratio test suggested that adding the random effect for Date made no difference at all to the model fit.

As such we are going for the model, model_1, without adding the random effect for Date


```{r Beta regression: are there significant difference in digestbility between feed}

null_model <- glmmTMB(DigestibilityCoeff ~ 1,
                       data = Samples,
                       family = beta_family(link = "logit"))

lrtest(model_1, null_model)

anova(null_model, model_1) # its the same thing as lrtest

```
since p = 1.09e-06 (p<0.05), this confirms that there is a significant effect of treatment somewhere among the groups.

## beta regression follow up: Pairwise comparisons 

to see which specific treatments differ from each other


```{r Pairwise comparison}

library(emmeans)
results = emmeans(model_1, pairwise ~ Sample.Name, type = "response")

# estimated mean digestibility per treatment

results$emmeans

# This gives us the back-transformed (proportion-scale) estimated digestibility coefficient for each treatment, with 95% confidence intervals:

# Response

# Extract the contrasts part and convert to a data frame
contrasts_df <- as.data.frame(results$contrasts)

# View it
contrasts_df

```
```{r plots}

library(ggplot2)
library(ggpubr)   # for stat_pvalue_manual
library(dplyr)
library(tidyr)

# 1. Prepare overall model p-value (from LRT)
p_values <- data.frame(
  label = paste0("p = ", signif(1.09e-06, 3))
)

# 2. Prepare pairwise contrasts for plotting
# Split "contrast" column (e.g. "1902 / 1905") into group1/group2
pairwise_p_values <- contrasts_df %>%
  separate(contrast, into = c("group1", "group2"), sep = " / ") %>%
  filter(p.value < 0.05) %>%   # keep only significant comparisons to avoid clutter
  mutate(
    label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  )

# 3. Set y.position for brackets (stack them above the data)
max_y <- max(Samples$DigestibilityCoeff, na.rm = TRUE)
pairwise_p_values$y.position <- seq(
  from = max_y * 1.5,
  by = max_y * 0.08,   # spacing between stacked brackets
  length.out = nrow(pairwise_p_values)
)

# 4. Plot
p <- ggplot(
  Samples,
  aes(x = Sample.Name, y = DigestibilityCoeff)
) +
  geom_boxplot(
    width = 0.6,
    aes(colour = Sample.Name)
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 20,
    size = 2,
    colour = "black",
    aes(group = Sample.Name)
  ) +

  # Overall model p-value
  geom_text(
    data = p_values,
    aes(x = Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = 1.1,
    vjust = 1.5,
    fontface = "italic",
    colour = "black"
  ) +

  # Pairwise significant comparisons
  stat_pvalue_manual(
    pairwise_p_values,
    label = "label",
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.4,
    size = 3.5,
    inherit.aes = FALSE
  ) +

  coord_cartesian(clip = "off") +
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.18))
  ) +
  theme_bw() +
  labs(
    x = "Ingredients",
    y = "Digestibility Coefficient",
    title = "Effect of ingredient on digestibility"
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "plain"),
    axis.text.x = element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1
    ),
    plot.margin = unit(c(1, 1, 1, 1), units = "cm")
  ) + 
  theme(legend.position = "none")

p



```


## Kruskal-Wallis test
The Kruskal-Wallis test (also called H-test) is a non-parametric statistical test used to compare three or more independent groups to determine whether their distributions differ. The Kruskal-Wallis test extends the Mann-Whitney U test to more than two independent groups.

The Kruskal-Wallis test is used when the assumptions for a one-way analysis of variance are not met. Because the Kruskal-Wallis test is a non-parametric test, the data used do not have to be normally distributed.

The data must at least have an ordinal scale level and the samples must be independent.

```{r statistics}

# residuals significantly deviate from normality

Samples$log_DC <- log(Samples$DigestibilityCoeff)
res.aov <- aov(log_DC ~ Sample.Name, data = Samples) #ANOVA
plot(res.aov, 2) #QQplot
aov_residuals <- residuals(object = res.aov) #extract ANOVA residuals
shapiro.test(x = aov_residuals) #run shapiro-wilk test

Samples$sqrt_DC <- sqrt(Samples$DigestibilityCoeff)
res.aov <- aov(sqrt_DC ~ Sample.Name, data = Samples) #ANOVA
aov_residuals <- residuals(object = res.aov) #extract ANOVA residuals
shapiro.test(x = aov_residuals) #run shapiro-wilk test


# or for proportions/percentages bounded 0-1:


summary(res.aov) #ANOVA results
b <- summary(res.aov)  

nmc <- pairwise.t.test(MHdig$Digestibility, MHdig$Ingredient, p.adjust.method = "BH") #t-test with no correction
nmc

d <- ggplot(MHdig, aes(x=Ingredient, y=Digestibility)) + 
  ylim(0,100)+
  geom_boxplot(width = 0.6, aes(colour = Ingredient)) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black", 
               position = position_dodge(width = 0.75)) +  # Adding mean points
  theme_bw()+ 
  ylab("Digestibility (%)")+
  xlab("Diet") +
  ggtitle("Crude Protein digestibility") +
  theme(axis.text = element_text(size = 10), 
        axis.text.x = element_text(vjust=0.3, ), 
        axis.title = element_text(size = 12, face = "plain"),                     
#        panel.grid.major.y = element_line(color = "gray", size = 0.75, linetype = 2),   # Adds background lines               
        plot.margin = unit(c(1,1,1,1), units = , "cm"))+  # Adding a margin
  annotate(geom="text", x=2.5, y=95, label="p = 5.25e-09 ***", fontface = "italic")+ #adds pvalue text to plot
  annotate(geom="text", x=1, y=80, label="a") + #adds pvalue text to plot
  annotate(geom="text", x=2, y=75, label="b") +#adds pvalue text to plot
  annotate(geom="text", x=3, y=77, label="b") +#adds pvalue text to plot
  annotate(geom="text", x=4, y=40, label="c") #adds pvalue text to plot


=
d
ggsave("MHdig_boxplot.png", width = 5, height = 5, dpi = 300)


```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
